Post-Disturbance Tree Community Trajectories in a Neotropical Forest

### 2018-07-03 WholePlotTrajectories.tex
graphs and relecture

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global changing context. The huge
diversity of tropical forests is often assumed to be shaped by constant
regime of disturbance yielding a diversity peak at intermediate
intensity, but for tropical forests this intermediate disturbance
hypothesis (IDH) remains debated, and this controversy also questions
the extent of communities resilience regarding their functional and
taxonomic facets is also to be clarified. To disentangle the ecological
processes driving communities response to disturbance we analysed the
taxonomic and functional diversity trajectories following a logging and
thinning disturbance gradient. Specifically we examined over 30 years
the trajectories of communities taxonomic richness and evennes and
functional composition, diversity and redundancy based on 7 leaf, stem
and life history traits. Trajectories highlighted the cyclic recovery of
communities taxonomic composition, maintaining the pre-disturbance
differences, as well as functional composition and diversity. The IDH
predicted communities functional response to disturbance that resulted
from competition processes determined by the disturbance intensity. The
IDH however poorly represented the taxonomic response that relied upon
lottery recruitment and recruitment limitation which hampered the
recovery of infrequent, shade tolerant species. After disturbance
communities functioning followed a cyclic recovery and the initial
composition differences among plots were maintained. Although consistent
the recovery of communities functioning and taxonomic composition and
diversity remained unachieved after 30 years which acknowledged the need
of decades-long recovery cycles to ensure a complete recovery, and
questioned communities sustainability after repeated disturbance.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local climate, and the carbon, water and nutrient cycles,
and ensure cultural and human well-being. The growing demand in forests
products together with current climatic changes increases the pressure
on remaining forests \citep{Gibson2011a, Morales-Hidalgo2015} and
threatens the maintenance and dynamics in space and time of communities
structure, composition and functioning
\citep{Anderson-Teixeira2013, Sist2015}.

In tropical forests, ecological communities are constantly re-shaped by
natural disturbance events that change both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions like competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{White2001, Chazdon2003a}. For now, this
has been largely studied through forest structural parameters, rapid and
convenient to measure, as aboveground biomass, tree height or stem
density \citep{Piponiot2016, Rutishauser2016}. These structural
parameters were thereafter sucessfully modeled, giving important
insights into the recovery of ecosystems processes and services
\citep{Herault2018}. However the response of forests diversity is still
unclear, albeit it determines the productivity, stability and
functioning of ecosystems \citep[\citet{Liang2016}]{Tilman2014} and
would be most probably impacted by the changes induced by disturbance
\citep{Baraloto2012a}. In the short-term, moderate disturbance may lead
to positive impacts on communities diversity, which have been formalized
by the intermediate disturbance hypothesis (IDH) stating a maximized
species diversity when the disturbance regime is moderate but
significant \citep{Molino2001, Kariuki2006a, Berry2008a}. Along time the
IDH would translate into a variety of post-disturbance trajectories
shaped by the different ecological processes at stake.

Validations of the IDH though remain scarce in the long term and mainly
rely on the analysis of species richness. Species richness though gives
limited or misleading information on forests recovery and functioning
\citep{Martin2015, Chaudhary2016}: more complete analysis would
encompass communities composition that is crucial for conservation
issues, and abundance distribution that reveals the ecological processes
involved \citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes and
assessing their role in the ecosystem would directly link communities
diversity to the ecosystem functioning and to its environmental
constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focused on major
traits related to species ecology and performance was sucessfully
adopted \citep{Diaz2005, Villeger2008a}. For example the functoinal
approach revealed in tropical rainforests the deterministic processes
fostering fast growing species with efficient resources acquisition
after disturbance \citep{Molino2001, Haddad2008, Ruger2009}. Disturbance
therefore entails a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. This
shift is translated into consistent trajectories of key functional
traits related to resource acquisition (leaf area, density and
chlorophyll content, and stem specific gravity and bark thickness), tree
growth and reproduction life history traits (seed mass and maximum
height) \citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}.
Eventually a complete overview of communities response to disturbance
would encompass the changes in functional redundancy, that quantifies
the amount of shared trait values among species \citep{Carmona2016}.
High functional redundancy like in the very diverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities
\citep{Trenbath1999, Elmqvist2003, Diaz2005}. High functional redundancy
also increases the functional overlap among species and would foster
neutral stochastic processes \citep{Gravel2006}.

To grasp all facets of communities response to disturbance we examined
here the taxonomic and functional trajectories in terms of richness,
evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. These trajectories would highlight
the recovery of communities initial characteristics and the underlying
ecological processes, and provide indications for future adaptive
conservation strategies \citep{Adler2007}. They would clarify the
tenants of the Intermediate Disturbance Hypothesis in the long term,
debated in tropical forests, and explicit the variety of
post-disturbance trajectories defined by the disturbance intensity. We
monitored over 30 years the response of 75 ha of neotropical forest
plots set up on a gradient of disturbance intensity, from 10 to 60\% of
ecosystem biomass removed. We made use of a large functional traits
database browsing major leaf, stem and seed functional traits and
species maximum height to draw the trajectories over time of communities
taxonomic and functional composition, richness, evenness and redundancy.
Specifically, we (i) questioned the recovery of communities taxonomic
and functional characteristics and identify the underlying assembly
processes, (ii) clarified the validity of the IDH in the long term for
tropical forest and elucidated its declination into different
trajectories, and (iii) questioned the length of complete communities
recovery.

@@

Paracou station in French Guiana (5°18'N and 52°53'W) is located in a
lowland tropical rain forest in a tropical wet climate with mean annual
temperature of 26°C, mean annual precipitation averaging 2980
mm.y\textsuperscript{-1} (30-y period) and a 3-month dry season
(\textless{} 100 mm.month\textsuperscript{-1}) from mid-August to
mid-November, and a one-month dry season in March \citep{Wagner2011}.
Elevation ranges between 5 and 50 m and soils correspond to thin
acrisols over a layer of transformed saprolite with low permeability
generating lateral drainage during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a
gradient of three logging, thinning and fuelwood cutting treatments
(Table \ref{tab:Tab1}) according to a randomized plot design with three
replicate blocks of four plots. The disturbance corresponds to averages
of 10 trees removed per hectare with a diameter at 1.3 m height (DBH)
above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm DBH for
treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3 (T3).
Treatments T2 and T3 besides included the thinning of trees by poison
girdling \citep{Schmitt1990, Blanc2009}. The disturbance intensity was
measured as the percentage of aboveground biomass (\%AGB) lost between
the first inventory in 1984 and five years after disturbance
\citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}.

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of Fabaceae, Chrysobalanaceae,
Lecythidaceae and Sapotaceae botanical families. In the twelve
experimental plots of the experiment, all trees above 10 cm DBH have
been mapped and measured annually since 1984. Trees are first identified
during inventories with a vernacular name assigned by the forest worker
team, and afterward with a scientific name assigned by botanists during
regular botanical campaigns. In 1984, specific vernacular names are
given to 62 commercial or common species whereas more infrequent ones
were identified under general identifiers only distinguishing trees and
palm trees. From 2003, botanical campaigns have been conducted every 5
to 6 years to identify all trees at the species level but identification
levels still varied among plots and campaigns.

This variability of protocols raised methodological issues as vernacular
names usually correspond to different botanical species. It resulted in
significant taxonomic uncertainties that had to be propagated to
composition and diversity metrics. The uncertainty propagation was done
through a Bayesian framework reconstituting complete inventories at
genus level from real incomplete ones on the basis of
vernacular/botanical names association. Vernacular names were replaced
through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,…, \alpha_3\big]\) observed across all
inventories between each vernacular name \emph{v} and the species
\(\big[s_1, s_2, …, s_N\big]\):

\begin{align}
M_v\Big(\big[s_1, s_2, …, s_N\big],\big[\alpha_1, \alpha_2,…, \alpha_3\big]\Big) \nonumber
\end{align}

@@

Six functional traits representing leaf economics (leaves thickness,
toughness, total chlorophyll content and specific leaf area, the leaf
area per unit dry mass) and wood economics (wood specific gravity and
bark thickness), and life history traits (maximum specific height and
seed mass) came from the BRIDGE project \footnote{http://www.ecofog.gf/Bridge/}.
Traits values were assessed from a selection of individuals located in
nine permanent plots in French Guiana, including two in Paracou, and
comprised 294 botanical species pertaining to 157 botanical genera.
Missing trait values were filled using multivariate imputation by
chained equation \citep{Mice2011}. Imputations were restricted within
genus, or family when samples were too scarce, in order to account for
the phylogenetic signal of the functional traits. Whenever a species
inventoried was not in the dataset, it was attributed a set of traits
values randomly sampled among species of the same next higher taxonomic
level (same genus or family). As seed mass information corresponds to a
classification into mass classes, no data filling process was applied
and analysis were restricted to the 414 botanical species recorded.

All composition and diversity metrics were obtained after 50 iterations
of the taxonomic uncertainty propagation framework and of the filling
process of missing trait values.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level,
\emph{i.e.} referring to the genus of observed or trialed botanical
names. Trajectories of communities taxonomic and functional variations
in composition after disturbance were followed in a two-dimensional NMDS
ordination space of the 30 years monitored. Two NMDS were conducted to
map either taxonomic flora inventories or communities functional
composition based on the 7 leaf, stem and life history traits (without
seed mass classes). In both cases the NMDS were performed using
abundance-based (Bray-Curtis) dissimilarity measures. Trajectories along
time were reported through the distance between successive inventories
and the reference inventories in 1989, \emph{i.e.} 2-3 years after
disturbance, when the uncertainty degree did not exceed 30\% of
undetermined trees. for the taxonomix composition trajectories were
measured by the euclidean distance between inventories in the
multivariate space of all inventories species. For the functoinal
composition trajectories were measured by the sum for all traits of the
euclidean distances between the traits density distributions of
successive inventories. To compensate the intrinsinc difference among
plots, the trajectories corresponded to the euclidean differences along
time with the reference inventory in 1989. Univariate trajectories of
the leaf, stem and life-history traits were also visualized with the
community weighted means (CWM) \citep{Diaz2007, Garnier2004}. Species
seed mass corresponded to 5 classes of increasing mass, seed mass
trajectories were therefore reported as the proportion of each class in
the inventories.

The taxonomic diversity was reported through species richness and the
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These two indices belong to the set of HCDT or
generalized entropy, respectively corresponding to the 0 and 2 order of
diversity (q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the Rao index of quadratic entropy which combines
species abundance distribution and average pairwise dissimilarity based
on all functional and life traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extremum of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB removed. The were
besides analysed through the linear correlations between Simpson and Rao
diversities and the initial \%AGB removed at 10, 20 and 30 years after
disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan from a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community: the functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

From 1989 (2-3 years after disturbance) to 2015 (28-29 years after
disturbance), 828388 individual trees and 591 botanical species
pertaining to 223 genus and 64 botanical families were recorded. For
undisturbed plots taxonomic Richness and Evenness (Simpson diversity)
remained stable over the 30 years monitored. In disturbed communities
after low disturbance intensity the taxonomic richness increased,
reaching a maximum gain of 14 botanical genera (plot 3 from treatment
2). After intense disturbance the taxonomic richness followed unimodal
trajectories, decreasing for ten years after disturbance before
recovering to pre-disturbance values. In all disturbed plots the
taxonomic evenness (Simpson diversity) followed unimodal trajectories
with first an increase until a maximum reached after around 20 years
that was positively correlated to the disturbance treatment
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). The return towards initial values
of communities evenness was beginning after 30 years except for two T3
plots (plots 8 and 12) which evenness still increased, suggesting
similar but delayed trajectories \ref{fig:DivTaxo}.

@@

The plot 7 from treatment 1 displayed a constantly outlying functional
evenness and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional evenness remained stable along the 30 years while in
disturbed plots it followed unimodal trajectories with a return towards
initial values starting around 20 years after disturbance.

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/FunDiv-1}

\caption{Trajectories of the Rao functional diversity over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals.}\label{fig:FunDiv}
\end{figure}

The impact of disturbance was examined specifically through the linear
correlation between the intial \%AGB removed and the Simpson and Rao
diversities (diversities of order 2) after 10, 20 and 30 years
\ref{fig:IDHplot}. The correlation with disturbance intensity was weak
for the Simpson diversity (\(R^2<0.25\)) and only valid from 20 years
after disturbance but it was much stronger for the Rao diversity
(\(0.60<R^2<0.75\)) for all the time studied. Slope of linear
correlations, reflecting the impact of disturbance, was the highest 20
years after disturbance.

@@

\caption{Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot1}
\end{figure*}\begin{figure*}

{\centering \includegraphics[width=1\linewidth]{WholePlotTrajectories_files/figure-latex/IDHplot-2}

\caption{Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot2}
\end{figure*}

@@

\subsubsection{Taxonomic and functional
trajectories}\label{taxonomic-and-functional-trajectories}

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
consistent trajectories over time after disturbance which revealed
significant compositional changes. According to the mapping of
functional traits (see appendix I) these compositional changes
corresponded to shifts towards species with more acquisitive functional
strategies, from communities with high average WD to high average SLA
and chlorophyll content. For disturbed communities the distance of
successive inventories to the 1989 reference inventory followed unimodal
trajectories translating cyclic compositional changes with a recovery of
the initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance treatment for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.91\) and
\(\rho_{spearman}^{functional}=0.96\) respectively) and the time at
maximum was reached around 26 years after disturbance for taxonomic
composition and 22 years for functional composition.

@@

\subsubsection{Traits community weighted means
(CWM)}\label{traits-community-weighted-means-cwm}

Changes in functional composition trajectories went hand to hand with
consistent trajectories of the 8 functional traits (Figure
\ref{fig:CWM}). Except for leaf chlorophyll content, which continued to
increase for some T3 and T2 plots 30 years after disturbance, all traits
and seed mass proportions followed unimodal trajectories either
stabilizing or returning towards their initial values. Thirty years
after disturbance the weighted means of communities specific maximum
height at adult stage (\emph{Hmax}), leaf toughness
(\emph{L\_toughness}) and wood specific gravity (\emph{WD}) remained
significantly lower than their initial value (Figure \ref{fig:CWM}). The
weighted means of bark thickness (\emph{Bark\_thick}) similarly remained
substantially higher than initially for all disturbed plots while the
specfic leaf area (\emph{SLA}) had almost recovered its initial value.
For all traits the maximum difference to initial state was correlated to
the disturbance intensity (\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\subsubsection{Functional redundancy}\label{functional-redundancy}

In control plots the functional redundancy remained stable but in
undisturbed plots it was quite variable and independent of the
disturbance intensity (See appendix I). Globally after most intense
disturbance (plots T2 and T3) communities redundancy decreased at first
place before increasing to edge, recover or exceed the initial value.

Considering the functional redundancy restricted to the functional space
of the initial inventory, all disturbed plots followed similar
decreasing humped shaped trajectories (@ref(fig:RedFun\_rest)). The
maximum redundancy loss was positively correlated with the disturbance
intensity (\(\rho_{spearman}=0.50\)) and the initial value had not
recovered for any disturbed communities.

@@

\subsection{Communities cyclic
recovery}\label{communities-cyclic-recovery}

Communities taxonomic and functional composition and functional evenness
proved resilient, following similar humped-shaped trajectories starting
to return towards pre-disturbance composition and functioning after 30
years.

>>>>

\Abstract{
Understanding the ecological rules underlying the maintenance of
tropical forests biodiversity, structure, functioning and dynamics is
urgent to anticipate their fate in the global change context. The huge
diversity of tropical forests is often assumed to be regularly reshaped
by natural disturbance yielding a diversity peak at intermediate
intensity, but this intermediate disturbance hypothesis (IDH) remains
debated, and this controversy also questions the extent of communities
resilience regarding their functional and taxonomic facets. To
disentangle the ecological processes driving community response to
disturbance, we analysed the taxonomic and functional diversity
trajectories following a disturbance gradient. Specifically, we
examined, over 30 years, the functional and taxonomic community
trajectories with regards to diversity, composition and redundancy.
Functional trajectories were drawn based on 7 leaf, stem and
life-history traits. We highlighted the cyclic recovery of community
taxonomic and functional composition. The pre-disturbance taxonomic
differences were maintained over time while the functional composition
trajectories were quite similar. The IDH did predict communities
functional diversity response while taxonomic diversity remained poorly
sensitive to disturbance intensity. Although consistent, the recovery of
community composition, diversity and redundancy remained unachieved
after 30 years. This acknowledged the need of decades-long cycles with
no disturbance to ensure a complete recovery, and questioned tropical
forest community resilience after repeated disturbance.
}

@@

The large areas covered with tropical forests worldwide hold crucial
environmental, economic and social values. They provide wood and
multiple non-timber forest products, shelter a diversified fauna,
regulate the local and regional climates, the carbon, water and nutrient
cycles, and ensure cultural and human well-being. The growing demand in
forests products together with current global changes increases the
pressure on remaining natural forests
\citep{Gibson2011a, Morales-Hidalgo2015} and threatens the maintenance
and dynamics in space and time of communities structure, composition and
functioning \citep{Anderson-Teixeira2013, Sist2015}.

In tropical forests, ecological communities are regularly re-shaped by
natural disturbance events that change both the abiotic environment,
through the fluxes of light, heat and water \citep{Goulamoussene2017},
and the biotic interactions such as competition among species
\citep{Chesson2000, Herault2018}. One of the cornerstone of tropical
forest ecology is to understand the processes and drivers of ecosystems
response to disturbance \citep{White2001, Chazdon2003a}. For now, this
has been largely studied through forest structural parameters such as
aboveground biomass, tree height or stem density
\citep{Piponiot2016, Rutishauser2016} that are rapid and convenient to
measure. These structural parameters have been sucessfully modeled,
giving important insights into the recovery of ecosystem processes and
services \citep{Herault2018}. However the response of forests diversity
and composition is still unclear, albeit it determines the productivity,
stability and functioning of ecosystems \citep{Tilman2014, Liang2016}.
In the short-term, moderate disturbance may lead to positive impacts on
communities diversity, an idea formalized by the intermediate
disturbance hypothesis (IDH) stating a maximized species diversity when
disturbance intensity is not too high
\citep{Molino2001, Kariuki2006a, Berry2008a}.

Validations of the IDH though remain scarce in the long-term and mainly
rely on the analysis of taxonomic richness \citep{Molino2001}. Taxonomic
richness, alone, gives limited or misleading information on forests
recovery and functioning \citep{Martin2015, Chaudhary2016}. More
ecological-meaningful analysis would couple richness to (i) evenness to
reveal the underlying ecological processes and (ii) composition that is
crucial for conservation issues
\citep{Magurran1988, Lavorel2002, Bellwood2006}. Furthermore, a
functional approach accounting for species biological attributes would
directly link communities diversity, composition and redundancy to
ecosystem functioning and to its environmental constraints
\citep{Violle2007b, Moretti2009, Baraloto2012a, Scheiter2013}. In that
respect, the functional trait-based approach that focused on major
traits related to species ecology and thta mediated species performance
in a given environment was sucessfully adopted
\citep{Diaz2005, Villeger2008a}. For instance, the functional approach
revealed in tropical rainforests the deterministic processes entailing,
after disturbance, a functional shift from a dominance of
``conservative'' slow-growing species dealing with scarce resources to
``acquisitive'' fast-growing species with rapid and efficient use of
abundant resources \citep{TerSteege2001, Reich2014, Herault2011}. This
shift is translated into consistent trajectories of key functional
traits related to resource acquisition (leaf and stem traits) and
life-history traits (seed mass, maximum size)
\citep{Wright2004, TerSteege2006, Westoby2006a, Chave2009b}. Eventually
a complete overview of communities response to disturbance would
encompass the changes in functional redundancy, that quantifies the
amount of shared trait values among species \citep{Carmona2016}. The
high functional redundancy of hyperdiverse tropical forests
\citep{Bellwood2006} mitigates the impacts of species removal on
ecosystem functioning and determines the resilience of communities after
disturbance \citep{Trenbath1999, Elmqvist2003, Diaz2005}.

In this study, we monitored over 30 years the response of 75 ha of
neotropical forest plots set up on a gradient of disturbance intensity,
from 10 to 60\% of ecosystem biomass removed. We made use of a large
functional traits database encompassing major leaf, stem and
life-history traits in order to draw the taxonomic and functional
trajectories in terms of richness, evenness, composition and redundancy
\citep{Lohbeck2015, Guariguata2001}. Specifically, we (i) questioned the
recovery of communities taxonomic and functional characteristics and
identify the underlying assembly processes, (ii) clarified the validity
of the IDH in the long term for tropical forest and elucidated its
declination into different trajectories, and (iii) questioned the length
of complete communities recovery.

@@

Paracou station in French Guiana (5??18'N and 52??53'W) is located in a
lowland tropical rain forest in a tropical wet climate with mean annual
temperature of 26??C, mean annual precipitation averaging 2980
mm.y\textsuperscript{-1} (30-y period) and a 3-month dry season
(\textless{} 100 mm.month\textsuperscript{-1}) from mid-August to
mid-November, and a one-month dry season in March \citep{Wagner2011}.
Elevation ranges between 5 and 50 m and soils correspond to thin
acrisols over a layer of transformed saprolite with low permeability
generating lateral drainage during heavy rains.

The experiment is a network of twelve 6.25ha plots that underwent a
disturbance gradient of three logging, thinning and fuelwood cutting
treatments (Table \ref{tab:Tab1}) according to a randomized plot design
with three replicate blocks of four plots. The disturbance corresponds
to averages of 10 trees removed per hectare with a diameter at 1.3 m
height (DBH) above 50 cm for treatment 1 (T1), 32 trees/ha above 40 cm
DBH for treatment 2 (T2) and 40 trees/ha above 40 cm DBH for treatment 3
(T3). Treatments T2 and T3 besides included the thinning of trees by
poison girdling \citep{Schmitt1990, Blanc2009}. The disturbance
intensity was measured as the percentage of aboveground biomass (\%AGB)
lost between the first inventory in 1984 and five years after
disturbance \citep{Piponiot2016} estimated with the BIOMASS R package
\citep{Biomass2018}.

@@

The study site corresponds to a tropical rainforest typical of the
Guiana Shield with a dominance of Fabaceae, Chrysobalanaceae,
Lecythidaceae and Sapotaceae. In the twelve experimental plots, all
trees above 10 cm DBH have been mapped and measured annually since 1984.
Trees are first identified with a vernacular name assigned by the forest
worker team, and afterward with a scientific name assigned by botanists
during regular botanical campaigns. In 1984, specific vernacular names
are given to 62 commercial or common species whereas more infrequent
ones were identified under general identifiers only distinguishing trees
and palm trees. From 2003, botanical campaigns have been conducted every
5 to 6 years to identify all trees at the species level but
identification levels still varied among plots and campaigns.

This variability of protocols in time raised methodological issues as
vernacular names usually correspond to different botanical species. It
resulted in significant taxonomic uncertainties that had to be
propagated to composition and diversity metrics. The uncertainty
propagation was done through a Bayesian framework reconstituting
complete inventories at genus level from real incomplete ones on the
basis of vernacular/botanical names association. Vernacular names were
replaced through multinomial trials based on the association probability
\(\big[\alpha_1, \alpha_2,???, \alpha_3\big]\) observed across all
inventories between each vernacular name \emph{v} and the species
\(\big[s_1, s_2, ???, s_N\big]\):

\begin{align}
M_v\Big(\big[s_1, s_2, ???, s_N\big],\big[\alpha_1, \alpha_2,???, \alpha_3\big]\Big) \nonumber
\end{align}

@@

Six functional traits representing leaf economics (leaves thickness,
toughness, total chlorophyll content and specific leaf area) and wood
economics (wood specific gravity and bark thickness), and life-history
traits (maximum specific height and seed mass) came from the BRIDGE
project \footnote{http://www.ecofog.gf/Bridge/}. Trait values were
assessed from a selection of individuals located in nine permanent plots
in French Guiana, including two in Paracou, and comprised 294 species
pertaining to 157 genera. Missing trait values were filled using
multivariate imputation by chained equation \citep{Mice2011}.
Imputations were restricted within genus, or family when samples were
too scarce, in order to account for the phylogenetic signal. Whenever a
species was not in the dataset, it was attributed a set of traits values
randomly sampled among species of the next higher taxonomic level (same
genus or family). As seed mass information was classified into classes,
no data filling process was applied and analyses were restricted to the
414 botanical species recorded.

All composition and diversity metrics were obtained after 50 iterations
of the taxonomic uncertainty propagation framework.

@@

To counter taxonomic uncertainties due to the variability of botanical
identification levels (in space) and protocols (in time), the taxonomic
composition and diversity analysis were conducted at the genus level.
Taxonomic and functional trajectories of community composition were
followed in a two-dimensional NMDS ordination space. Two NMDS were
conducted to map either taxonomic or functional composition, the later
based on the 7 leaf, stem and life history traits (without seed mass
classes). NMDS were performed using abundance-based (Bray-Curtis)
dissimilarity measures. Trajectories along time were reported through
the euclidean distance between the target inventories and the reference
inventories in 1989, \emph{i.e.} 2-3 years after disturbance. Univariate
trajectories of the leaf, stem and life-history traits were also
visualized with the community weighted means (CWM)
\citep{Diaz2007, Garnier2004}. Species seed mass corresponded to 5
classes of increasing mass, seed mass trajectories were therefore
reported as the proportion of each class in the inventories (Supp Mat
XX).

The taxonomic diversity was reported through species richness and the
evenness, \emph{i.e} the Hill number translation of the Simpson index
\citep{Hill1973}. These two indices belong to the set of HCDT or
generalized entropy, respectively corresponding to the 0 and 2 order of
diversity (q), recomended for diversity studies
\citep{Patil1982, Tothmeresz1995, Marcon2015}. The functional diversity
was reported using the Rao index of quadratic entropy which combines
species abundance distribution and average pairwise dissimilarity based
on all functional traits.

The impacts of initial disturbance were tested with the spearman rank
correlation between the extrema of taxonomic and functional metrics
reached over the 30 years and the initial \%AGB removed. The were
besides analysed through the linear correlations between Simpson and Rao
diversities and the initial \%AGB removed at 10, 20 and 30 years after
disturbance.

The functional redundancy was measured as the overlap among species in
community functional space \citep{Carmona2016}. The samples of the trait
database were first mapped in a 2-dimensional plan with a PCA analysis.
Then, multivariate kernel density estimator associated with individual
trees returned species traits probability distribution (TDP). Species
TDP weighted by species abundance were eventually summed for each
community: the functional redundancy was the sum of TDPs overlap,
expressed as the average number of species that could be removed from
without reducing the functional space (see appendix I for a more
comprehensive sheme).

@@

From 1989 (2 years after disturbance) to 2015 (28 years after
disturbance), 828388 individual trees and 591 botanical species
pertaining to 223 genus and 64 families were recorded. For undisturbed
plots, taxonomic Richness and Evenness remained stable over the 30 years
monitored. In disturbed communities, after low disturbance intensity the
taxonomic richness increased, reaching a maximum gain of 14 botanical
genera (plot 3 from treatment 2). After intense disturbance the
taxonomic richness followed a more complex trajectory, decreasing for
ten years after disturbance before recovering to pre-disturbance values.
In all disturbed plots the taxonomic evenness first increased until a
maximum reached after around 20 years. This maximum was positively
correlated to the disturbance intensity
(\(\rho_{spearman}^{Shannon}=0.86\), and
\(\rho_{spearman}^{Simpson}=0.89\)). The evenness then stabilized except
for two T3 plots (plots 8 and 12) for which evenness still increased.

@@

The plot 7 from treatment 1 displayed a constantly outlying functional
diversity and was removed from the graphical representation for better
readability (see appendix for full graphs). In undisturbed plots the
functional diversity remained stable along the 30 years. In disturbed
plots, trajectories depend on the disturbnace intensity with, for low
intensity, a low but long-lasting increase up to amximum reached after
20-25 years and, for, high intensity, a fast but short increase
followed, after 10 years, by a slow decrease towards the inital values.

The impact of disturbance intensity was examined through the linear
correlation between (i) the \%AGB removed and (ii) taxonomic richness,
evenness and functional diversity after 10, 20 and 30 years
\ref{fig:IDHplot}. The correlation with disturbance intensity was weak
for XXXXXXXXXX, for the taxonomic evenness (\(R^2<0.25\)) while highly
significant for functional diversity (\(0.60<R^2<0.75\)) whatever the
time since disturbance.

@@

\caption{Upper panels, Trajectories of the taxonomic evenness (Simpson diversity) \textbf{(a)} and Rao functional diversity \textbf{(b)} over 30 years after disturbance. Colors are treatments: green (control), blue (T1), orange (T2), red (T3) with shaded areas the credibility intervals. Lower panels, Relationship between the initial \%AGB removed and Simpson \textbf{(c)} and Rao \textbf{(d)} diversities 10, 20 and 30 years after disturbance.}\label{fig:IDHplot}
\end{figure*}

@@

While both taxonomic and functional composition remained stable in
undisturbed communities (Figure \ref{fig:NMDSplans}), they followed
marked and consistent trajectories over post-disturbance time. in
disturbed communities, these compositional changes corresponded to
shifts towards species with more acquisitive functional strategies, from
communities with high average WD to high average SLA and chlorophyll
content (see appendix I). For functional composition, this translated
into cyclic compositional changes with an unachieved recovery of the
initial composition (Figure \ref{fig:NMDSplans}). The maximum
dissimilarity with the initial state was positively correlated to the
disturbance intensity for both taxonomic and functional composition
(\(\rho_{spearman}^{taxonomic}=0.91\) and
\(\rho_{spearman}^{functional}=0.96\) respectively). The maximum value
was reached around 26 years after disturbance for taxonomic composition
and 22 years for functional composition.

@@

Except for leaf chlorophyll content, which continued to increase for
some T3 and T2 plots 30 years after disturbance, all traits and seed
mass proportions followed unimodal trajectories either stabilizing or
returning towards their initial values.

Maximum height at adult stage (\emph{Hmax}), leaf toughness
(\emph{L\_toughness}) and wood density (\emph{WD}) first decreased and
then slightly increased but remained significantly lower than their
initial value (Figure \ref{fig:CWM}). On the other side, Bark thickness
(\emph{Bark\_thick}) and specific leaf area (\emph{SLA}) increased and
while \emph{Bark\_thick} remained substantially high after 30 years,
\emph{SLA} had almost recovered its initial value. For all traits, the
maximum difference to initial value was correlated to the disturbance
intensity (\(\rho_{spearman}^{L_{thickness}}=0.67\),
\(\rho_{spearman}^{L_{chloro}}=0.45\),
\(\rho_{spearman}^{L_{toughness}}=-0.43\),
\(\rho_{spearman}^{SLA}=0.93\), \(\rho_{spearman}^{WD}=-0.78\),
\(\rho_{spearman}^{Bark-thickness}=0.88\),
\(\rho_{spearman}^{Hmax}=-0.48\)).

@@

\subsection{Functional redundancy}\label{functional-redundancy}

All disturbed plots had lower functional redundancy than control plots
and followed similar hump-shaped trajectories (@ref(fig:RedFun\_rest)).
The maximum redundancy loss was positively correlated with the
disturbance intensity (\(\rho_{spearman}=0.50\)) and the initial value
had not recovered for any disturbed communities after 30 years.

@@

\subsection{A cyclic recovery of community
composition}\label{a-cyclic-recovery-of-community-composition}
